---
title: "Lab 9 - Two-Way ANOVA"
author: "Sonora Meiling, adapted from Marilyn Brandt 2016"
date: "created 10/17/2022 for MES 503 Fall 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

*The goal of this lab is to introduce you to working with the functions necessary for performing two-way ANOVAs in R.*

### Housekeeping
By now you should be getting used to the housekeeping steps. Make sure you are working in a "Lab 9" project that is saved in an appropriately named and organized folder on your computer. 

As always, let's first define our working directory

```{r Define the Working Directory}
setwd("C:/Users/sonor/Desktop/MES 503/Fall 2022/Lab 9- Two-way ANOVA")

```

### Example 1: Effect of Nutrients on Coral Lesions
The first dataset uses the corallesions.csv file, and includes the number of lesions on corals exposed to different concentrations of Phosphorus and Nitrogen. Two-way ANOVAs are similar to one-way ANOVAs in many ways, use your previous one-way ANOVA examples as references.

```{r Import the Data}
CoralLesions <- read.csv("data/corallesions.csv")
```

What are the independent and dependent variables? Specify if each dependent variable is random or fixed

    - Independent: Phosphorus and nitrogen concentration
    - Dependent: number of lesions

What are the statistial null and alternative hypotheses for the ANOVA test?

     - Null: mean number of lesions on corals at all phosphorous levels are equal; mean number of lesions on corals at all nitrogen levels are equal; response is consistent across all phosphorous and nitrogen level combinations 
     - Alternative: mean number of lesions on corals at all phosphorous levels are not equal; mean number of lesions on corals at all nitrogen levels are not equal; response is not consistent across all phosphorous and nitrogen level combinations

What are the research null and alternative hypotheses for the ANOVA test?

     - Null: phosphorous has no effect on lesion abundance; nitrogen has no effect on lesion abundance; the combination of phosporous and nitrogen has no effect on coral lesion abundance
     - Alternative: phosphorous has an effect on lesion abundance; nitrogen has an effect on lesion abundance; the combination of phosporous and nitrogen has an effect on coral lesion abundance

```{r Look at the data}
#### Dataframe ####
View(CoralLesions)

# Let's get rid of the "Sample" column because it is unnecessary
CoralLesions$Sample <- NULL
colnames(CoralLesions)[3] <- c("Lesions")

# Do we need to rearrange (e.g. melt) the data? If so, do so
# no :)

#### Plot ####
##### Summarize the Data #####
# In order to make bar plots of our data, we first have to summarize the data for ggplot

library(plyr)

# The format for making the summary dataframe is below, words in all caps are directions for what you should input to customize the command to your data. Note the addition of the second factor since we are running a two-way ANOVA
# DATAFRAME.summary <- ddply(DATAFRAME, ~FACTOR1 * FACTOR2, summarise, 
                    #mean = mean(RESPONSEVARIABLE), 
                    #sd = sd(RESPONSEVARIABLE), 
                    #n = length(RESPONSEVARIABLE), 
                    #SEM = sd(RESPONSEVARIABLE)/sqrt(length(RESPONSEVARIABLE)))

CoralLesions_summary <-ddply(CoralLesions, ~P*N, summarise,
                          mean = mean(Lesions),
                          sd = sd(Lesions),
                          n = length(Lesions),
                          SEM = sd(Lesions)/sqrt(length(Lesions)))
#This can be done with the formula (y ~ A + B + A:B) or (y ~ A*B)
CoralLesions_summary

# Now we can actually plot the data using the following skeleton, again, words in all caps are directions for what you should input to cusotmize the command to your data
#ggplot(aes(x = FACTOR1, y = mean, fill = FACTOR2), data = DATAFRAME) +
  #geom_bar(stat = "identity", position = position_dodge()) +
  #geom_errorbar(aes(ymin = mean - SEM, ymax = mean + SEM), width = .25, position = position_dodge()) +
  #ggtitle("TITLE OF YOUR FIGURE") +
  #ylab("Y AXIS TITLE") +
  #xlab("X AXIS TITLE") + 
  #scale_fill_discrete(name = "FACTOR2")

library(ggplot2)
CoralLesion_Plot <- ggplot(aes(x = P, y = mean, fill = N), data = CoralLesions_summary) + #by using the "fill" argument we are able to plot the different concentrations of N as different bars in groups within each concentration of P on the x axis + 
  geom_bar(stat="identity", position = position_dodge()) + # You have to dodge the bars so they don't get stacked on top of each other
  geom_errorbar(aes(ymin = mean - SEM,ymax = mean + SEM), position = position_dodge(width = 0.9), width = 0.25) + # You also have to dodge the errorbars
  ylab("Number of Lesions (mean +/- SEM)") + xlab("Phosphorus concentration") + scale_fill_discrete(name = "Nitrogen concentration") +
  ggtitle("Effect of Nutrient Enrichment on Coral Disease")

CoralLesion_Plot
```

Does it look like there is a diffrence in coral lesions among phosphorous concentrations? Across nitrogen concentrations? Across the interaction of the two nutrients?


```{r Test Assumptions}
##### Assumption 1: Random Samples #####
# This assumption is satisfied by the experimental design

##### Assumption 2: Independent Samples #####
# This assumption is satisfied by the experimental design

##### Assumption 3: Equal Variance #####
#Here is where two-way ANOVAs begin to differ. Since there are two factors, we have to test the variance among each of the 9 groups (we can easily create these groups as vectors using the split() command). 
treatment_groups <- split(CoralLesions$Lesions,list(CoralLesions$P, CoralLesions$N))
View(treatment_groups)

#now run the bartlett test as we are used to from the one-way ANOVA
CoralLesions_bartlett <- bartlett.test(treatment_groups)
CoralLesions_bartlett


# What are the results? What do they mean? Do number of lesions in different treatments have equal variance? Do we need to transform the data?
# yes, so we don't need to transform the data :)


#### Assumption 4: Normal Distribution ####
#Now let's check the distribution of the residuals. Using the residuals() command we can easily extract the residuals from the ANOVA, so we first have to run and store the ANOVA

CoralLesions_ANOVA <- aov(Lesions ~ P*N, CoralLesions) #since there are now 2 factors, we have to analyze the response variable by factor A, factor B, and the interaction of factor A and B. Same as when we summarized the data, this can be done with the formula (y ~ A + B + A:B) or (y ~ A*B)
  
#now that we have the ANOVA, we can extract them using the residuals() function
CoralLesions_residuals <- residuals(CoralLesions_ANOVA)

#now we can easily plot the histogram to look at the distribution the same way as we did with the one-way ANOVAs
hist(CoralLesions_residuals)
#do the residuals look approximately normally distributed?

#we can statistically test this using the shapiro-wilk test the same way as we did with the one-way ANOVAs
shapiro.test(CoralLesions_residuals)

#do we need to transform the data? If so, do and create a new ANOVA object with the transformed data
# the residuals are not normally distributed which means we need to transform the data :(


lesions_sqrt <- sqrt(CoralLesions$Lesions)
lesions_sqrt
CoralLesions$lesions_sqrt <- c(lesions_sqrt)  # append the transformed data as a new column in your original coral lesions data frame.

#Next, we have to extract the new residuals to run the shapiro test
CoralLesions_sqrt_ANOVA <- aov(lesions_sqrt ~ P*N,CoralLesions)
CoralLesions_sqrt_residuals <- residuals(CoralLesions_sqrt_ANOVA)
#do they look more normally distributed?
hist(CoralLesions_sqrt_residuals)

#now rerun the shapiro test with the transformed data to see if our transformation reshapes the data to adhere with the ANOVA assumptions
shapiro.test(CoralLesions_sqrt_residuals)

# yes, the residuals are now normally distributed and this transformation worked 

#Let's make sure there isn't a better transformation before we move on.

lesions_squared <- (CoralLesions$Lesions)^2

CoralLesions$lesions_sqrd <- lesions_squared

#Next, we have to extract the new residuals to run the shapiro test
CoralLesions_sqrd_ANOVA <- aov(lesions_sqrd ~P*N, CoralLesions)
CoralLesions_sqrd_residuals <- residuals(CoralLesions_sqrd_ANOVA)
#do they look more normally distributed?
hist(CoralLesions_sqrd_residuals)

#now rerun the shapiro test with the transformed data to see if our transformation reshapes the data to adhere with the ANOVA assumptions
shapiro.test(CoralLesions_sqrd_residuals)

#Use the p-value to determine if this is a better transformation than the square-rooted data

# yes, the residuals are now normally distributed so this transformation worked :) This transformation has a higher p-value (0.2419) so we will use this transformation
```

Now that we know we meet all of the assumptions of an ANOVA, we can run the analsis.

```{r Results}
# Remember we have already built the ANOVA, now we just need to look at the results
CoralLesions_sqrd_ANOVA
summary(CoralLesions_sqrd_ANOVA)
```

What are the results of the experiment in respect to the effects of nutrient enrichment on coral disease? HINT: Make sure to address all 3 hypotheses and include the F-statistic and p-value

      - There was no significant difference in mean coral lesion abundance across the phosphorous and nitrogen combinations (F = 2.090, p = 0.0979). -or- There was no significant effect of the interaction of phosphorous and nitrogen on mean coral lesion abundance (F = 2.090, p = 0.0979)
      - There was a signficant effect of phosphorus on mean coral lesion abundance (F = 7.826, p = 0.001).
      - There was a signficant effect of nitrogen on mean coral lesion abundance (F = 5.152, p = 0.010)
      
```{r Plot the Results}
#The plot we made previously focuses on displaying the interaction of P and N on lesions, however, since the interaction of P and N was not significant, it makes more sense to create two separate plots of each grouping. Our previous summary dataframe calculated parameters based on the interaction of P and N, so if we want to plot just one of the groupings, we need to create a new summary dataframe
CoralLesions_Psummary <- ddply(CoralLesions, ~P, summarise, 
                            mean = mean(Lesions),
                            sd = sd(Lesions),
                            n = length(Lesions),
                            SEM = sd(Lesions)/sqrt(length(Lesions)))
CoralLesions_Psummary

#now we can plot the number of lesions within each treatment separately
CoralLesions_PPlot <-  ggplot(aes(x = P, y = mean), data = CoralLesions_Psummary) + 
  geom_bar(stat="identity", position = position_dodge()) + 
  geom_errorbar(aes(ymin = mean - SEM,ymax = mean + SEM), position = position_dodge(width = 0.9), width = 0.25) + 
  ylab("Number of Lesions (mean +/- SEM)") + 
  xlab("Phosphorus Concentration") +
  ggtitle("Effect of Phosphorous on Coral Disease")
CoralLesions_PPlot


#reordering the p x axis in increasing levels of phosphorous 
phosphorous_order <- factor(CoralLesions_Psummary$P, level = c("Ambient", "10x", "20x"))

PPlot_reordered <- ggplot(aes(x = phosphorous_order, y = mean), data = CoralLesions_Psummary) + 
  geom_bar(stat="identity", position = position_dodge()) + 
  geom_errorbar(aes(ymin = mean - SEM,ymax = mean + SEM), position = position_dodge(width = 0.9), width = 0.25) + 
  ylab("Number of Lesions (mean +/- SEM)") + 
  xlab("Phosphorus Concentration") +
  ggtitle("Effect of Phosphorous on Coral Disease")

#Now create the plot for the effect of Nitrogen on coral disease
CoralLesions_Nsummary <- ddply(CoralLesions, ~N, summarise, 
                            mean = mean(Lesions),
                            sd = sd(Lesions),
                            n = length(Lesions),
                            SEM = sd(Lesions)/sqrt(length(Lesions)))
CoralLesions_Nsummary

#now we can plot the number of lesions within each treatment separately
CoralLesions_NPlot <-  ggplot(aes(x = N, y = mean), data = CoralLesions_Nsummary) + 
  geom_bar(stat="identity", position = position_dodge()) + 
  geom_errorbar(aes(ymin = mean - SEM,ymax = mean + SEM), position = position_dodge(width = 0.9), width = 0.25) + 
  ylab("Number of Lesions (mean +/- SEM)") + 
  xlab("Nitrogen Concentration") +
  ggtitle("Effect of Nitrogen on Coral Disease")
CoralLesions_NPlot

```

But we're not done yet! Remember, so far we only know that there is a significant difference in number of lesions among P and N concentrations, but not which are significantly different from each other.

```{r Tukey's HSD Post-hoc analysis}
library(agricolae)
CoralLesions_Ptukey <- HSD.test(CoralLesions_sqrd_ANOVA, "P")
CoralLesions_Ptukey

#What are the results? Which treatment has the most lesions? The least? Which treatments are significantly different from each other?


#add the significant letters to the barchart
P_sigletters <- c("b", "ab", "a")
CoralLesions_Psummary$sigletters <- P_sigletters

PPlot_reordered + 
  geom_text(aes(x = phosphorous_order, y = mean+SEM + 0.25), data = CoralLesions_Psummary, label = P_sigletters, colour = "black")

# There was a signficant effect of phosphorus on mean coral lesion abundance (F = 7.826, p = 0.001). Corals exposed to 20x phosphorous had the most lesions, while corals exposed to ambient levels of phorphorous had the least lesions. 20x levels of phosphorous significantly increased the abundance of coral lesions (Tukey HSD post-hoc analysis p < 0.05).



#repeat for the nitrogen treatment
CoralLesions_Ntukey <- HSD.test(CoralLesions_sqrd_ANOVA, "N")
CoralLesions_Ntukey

#What are the results? Which treatment has the most lesions? The least? Which treatments are significantly different from each other?


#add the significant letters to the barchart
N_sigletters <- c("ab", "a", "b")
CoralLesions_Nsummary$Nsigletters <- N_sigletters

CoralLesions_NPlot + 
  geom_text(aes(x = N, y = mean+SEM + 0.25), data = CoralLesions_Nsummary, label = N_sigletters, colour = "black")

#What are the results? Which treatment has the most lesions? The least? Which treatments are significantly different from each other?

# There was a signficant effect of nitrogen on mean coral lesion abundance (F = 5.152, p = 0.010). Corals exposed to 20x nitrogen had the most lesions, while corals exposed to ambient levels of nitrogen had the least lesions. 20x levels of nitrogen significantly increased the abundance of coral lesions (Tukey HSD post-hoc analysis p < 0.05).

```

### ANOVA Steps
To review, the steps of performing an ANOVA are
   1. Look at the data (data frame, plot/ggplot)
   2. Test for equal variances (bartlett.test())
   3. Transform if necessary
   4. Save residuals and check for normality (shapiro.test())
   5. Transform if necessary 
   6. Perform the ANOVA (aov())
   7. Perform post-hoc comparisons if necessary

### Example 2: Millepora Distribution
The second dataset usess millep.csv, and includes densities of Millepora coral colonies in 3 habitats at 3 sites in Panama.

```{r Import the Data}
millepora <- read.csv("data/millepora.csv")
```

What are the independent and dependent variables? Specify if each dependent variable is random or fixed

    - Independent: habitat and site
    - Dependent: millepora density

What are the statistial null and alternative hypotheses for the ANOVA test?

     - Null:  mean density of millepora at all habitat are equal; mean density of millepora at all sites are equal; response is consistent across all habitat and site combinations
     - Alternative: mean density of millepora at all habitat are not equal; mean density of millepora at all sites are not equal; response is not consistent across all habitat and site combinations

What are the research null and alternative hypotheses for the ANOVA test?

     - Null: there is no effect of habitat on millepora density, there is no effect of site on millepora density, there is no effect of the interaction of habitat and site on millepora density
     - Alternative: there is an effect of habitat on millepora density, there is an effect of site on millepora density, there is an effect of the interaction of habitat and site on millepora density

```{r Look at the data}
#### Dataframe ####
View(millepora)
# Do we need to rearrange (e.g. melt) the data? If so, do so
# no it is already tabular :)

#### Plot ####
##### Summarize the Data #####
# In order to make bar plots of our data, we first have to summarize the data for ggplot

millepora_summary <- ddply(millepora, ~Site*Habitat, summarise,
                          mean = mean(Density.of.Millepora),
                          sd = sd(Density.of.Millepora),
                          n = length(Density.of.Millepora),
                          SEM = sd(Density.of.Millepora)/sqrt(length(Density.of.Millepora)))

millepora_plot <- ggplot(aes(x = Site, y = mean, fill = Habitat), data = millepora_summary)  + 
  geom_bar(stat="identity", position = position_dodge()) + 
  geom_errorbar(aes(ymin = mean - SEM,ymax = mean + SEM), position = position_dodge(width = 0.9), width = 0.25) + 
  ylab("Millepora Density (mean +/- SEM)") + 
  xlab("Site") + 
  scale_fill_discrete(name = "Habitat") +
  ggtitle("Millepora Density")
millepora_plot

```

Does it look like there is a diffrence in millepora density among habitats ? Among sites? Among the interaction of habitat and site?
Millepora density looks pretty consitent across habitat and site

```{r Test Assumptions}
##### Assumption 1: Random Samples #####
# This assumption is satisfied by the experimental design

##### Assumption 2: Independent Samples #####
# This assumption is satisfied by the experimental design

##### Assumption 3: Equal Variance #####
millepora_groups <- split(millepora$Density.of.Millepora,list(millepora$Site, millepora$Habitat))

#now run the bartlett test
millepora_bartlett <- bartlett.test(millepora_groups)
millepora_bartlett

# Do we need to transform the data?
# yes, we have equal variance so we do not need to transform the data :)


#### Assumption 4: Normal Distribution ####
#Now let's check the distribution of the residuals. Using the residuals() command we can easily extract the residuals from the ANOVA, so we first have to run and store the ANOVA
millepora_ANOVA <- aov(Density.of.Millepora ~ Site*Habitat, millepora)

millepora_residuals <- residuals(millepora_ANOVA)

hist(millepora_residuals)

shapiro.test(millepora_residuals)
# Do we need to transform the data? If so, do
# yes, the residuals are normally distributed and we don't need to transform the data :)
```

Now that we know we meet all of the assumptions of an ANOVA, we can run the analsis.

```{r Results}
# Remember we have already built the ANOVA, now we just need to look at the results
summary(millepora_ANOVA)
```

What are the results of the experiment in respect to the effects of site and habitat on millepora density? HINT: Make sure to address all 3 hypotheses and include the F-statistic and p-value

      - There was a significant effect of the interaction of site and habitat on millepora denisty (F = 6.181, p = 0.001)

```{r Tukey's HSD Post-hoc analysis}
#Do we need to run post-hoc analyses? If so, do so and visualize the results on a revised bar chart(s)
interaction_tukey <- HSD.test(millepora_ANOVA,"Site"*"Habitat")

#returns a null because HSD.test cannot handle interaction terms
#we can use the TukeyHSD command and manually asign letter groups based on the p-values

interaction_tukey2 <- TukeyHSD(millepora_ANOVA)
interaction_tukey2
#a: Uva, Iguanas
#b: Secas, Iguana

#Uva a
#Secas b
#Iguana ab

#but this is a lot of work, especially if you have a lot of levels within factor(s)
#so another way to do this is to use glm modeling
library(lme4)
install.packages("multcomp")
library(multcomp)
millepora$factorAB <- with(millepora, interaction(Site, Habitat)) #This adds the interaction term as another column to the original dataframe
View(millepora)

# General linear model 
glm.mod <- glm(Density.of.Millepora ~ Site + Habitat + factorAB, family = 'gaussian', data = millepora) 

# Now we can use our favorite HSD.test to run the multiple comparisons test

millepora_tukey <- HSD.test(glm.mod, trt = "factorAB", group = TRUE)
millepora_tukey

#What are the results? Which site:habitat combo has the most lesions? The least? Which locations are significantly different from each other?


interaction.sigs <- c("ab","b","ab","b","b","b", "ab","a","b")
millepora_summary$interaction.sigs <- interaction.sigs
millepora_plot + geom_text(aes(x = Site, y = mean+SEM + 1.3), data = millepora_summary, label = interaction.sigs, position = position_dodge(width=0.9), colour = "black")# position dodge makes is so that the letters aren't overlapping eachother, will have to adjust the width for each figure to properly fit them

#There was a significant effect of the interaction of site and habitat on millepora denisty (F = 6.181, p = 0.001). Millepora density is significantly greater at Uva mid-depth than Uva shallow, all habitats at Secas, and Iguana mid-depth (Tukey HSD post-hoc analysis p < 0.05). Millepora density was consistent across all habitats at Secas and Iguana (Tukey HSD post-hoc analysis p < 0.05).

library(TukeyC)
millepora_posthocT = TukeyC(millepora_ANOVA, which='Site')

```

